### 2021-11-23 text/outline/draft.tex
add decrease ep_tol in Gaussian expt to 1e-7
add some simulation text

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A class of exponential family measurement error models, with application to single-cell CRISPR screens}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``exponential family errors-in-variables model''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we conduct a careful study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method. Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors. Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

\subsection{Model}
\subsubsection*{Negative binomial model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
\end{equation}
\begin{equation}\label{glmeiv_model_2}
\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_3}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
\end{equation}
\begin{equation}\label{glmeiv_model_4}
\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as
\begin{equation}\label{glmeiv_model_5}
p_i \sim \textrm{Bern}(\pi),
\end{equation} where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.

@@

\subsection{Computation}
We develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data. Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure) or high-performance cluster.

@@

@@

\section{Additional simulation results}

>>>>

\begin{document}
\noindent
Tim, Gene, Kathryn
\begin{center} 
\textbf{A class of exponential family measurement error models applied to single-cell CRISPR screens}
\end{center}

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method. Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure, Amazon Web Services) and high-performance clusters. Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
\end{itemize}
The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration. The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, controlling for the technical factors. Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).

@@

\subsection{Model}\label{sec:model}
\subsubsection*{Negative binomial model}
Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})). We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities. To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed. We model the gene expression counts according to
\begin{equation}\label{glmeiv_model_1}
m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
\end{equation}
\begin{equation}\label{glmeiv_model_2}
\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
\end{equation}
where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants. The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$. Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size). The model for the gRNA counts is 
\begin{equation}\label{glmeiv_model_3}
g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
\end{equation}
\begin{equation}\label{glmeiv_model_4}
\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
\end{equation}
where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants. We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}. Finally, we model the marginal perturbation probability as
\begin{equation}\label{glmeiv_model_5}
p_i \sim \textrm{Bern}(\pi),
\end{equation} where $\pi \in (0,1/2]$. 
 Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model. The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model. Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.

@@

\subsection{Computation}
We develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data. Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.

@@

We conducted a simulation study to compare the empirical performance of GLM-EIV and the thresholding method. We generated data on $n = 150,000$ cells from the GLM-EIV model (Section \ref{sec:model}), setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$. Note that $\beta^m_1 = \log(0.25)$ denotes a reduction in gene expression by a factor of $4$; this is a fairly large effect size on the order of a positive control pair. We included ``sequencing batch'' (Bernoulli-distributed) as a covariate and sequencing depth (Poisson-distributed) as an offset. We varied the log-fold change in gRNA expression $\beta^g_1$ over a grid defined on $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problem settings. Finally, we generated the gene expression and gRNA count data using two response distributions: Poisson and negative binomial (with fixed size parameter $\theta = 20$). For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ datasets. Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.

We applied three methods to the simulated data: ``standard'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression. We applied the methods to the negative binomial data twice: once taking the size parameter $\theta$ as known, and once taking $\theta$ as unknown. In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}. We display the results of the simulation study in Figure \ref{main_text_sim}. Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), and rows correspond to performance metrics (i.e., bias, mean-squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time). The parameter $\beta^g_1$ is plotted on the $x$-axis. The methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).

First, we observed that GLM-EIV uniformly dominated thresholded regression on statistical performance metrics. GLM-EIV exhibited lower bias (row 1) and mean-squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4). Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments. Additionally, we observed that accelerated GLM-EIV exhibited identical performance to (standard) GLM-EIV on statistical performance metrics (rows 1-4) despite having a substantially reduced computational cost (bottom row). In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).

% Interestingly, thresholded regression exhibited better CI coverage for theta est. rather than theta known (row 3). Estimating theta leads to slight \textit{inflation} bias (i.e., \textit{overestimating} the effect size), whereas thresholdeding leads to slight \textit{attenuation} bias (as demonstrated previously). These two phenomena cancel, causing the thresholding method to exhibit better CI coverage when estimating theta. Our focus on this work is \textit{not} on negative binomial size parameter estimation; GLM-EIV and thresholded regression are (in principle) compatible with \textit{any} negative binomial size parameter estimation procedure, including those based on more sophisticated techniques, such as regularization. An interesting future direction would be to assess the performance of the methods under different size parameter estimation approaches.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
	\caption{\textbf{Simulation study}. Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time). Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels. GLM-EIV demonstrated superior statistical performance to the thresholding method across all metrics (rows 1-4). Accelerated GLM-EIV had substantially lower computational cost than standard GLM-EIV (bottom row) despite having exhibited identical statistical performance (rows 1-4).}
	\label{main_text_sim}
\end{figure}

@@

\section{Additional simulation results}\label{sec:extra_sims}
\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{../../figures/supplement_sim/gaussian}
	\caption{}
	\label{fig:gaussian_sim}
\end{figure}
